---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# arsampler

<!-- badges: start -->
<!-- badges: end -->

## Overview

`arsampler` is a package to help perform the acceptance-rejection sampling without having to code the entire algorithm itself. The current version of this package (Version 0.1.0) can only accept a proposal densities from pre-defined templates while the user can define their own target densities (see examples below).


## Installation

You can install the released version of `arsampler` along with the unit tests with:

``` r
# Install the relevant library (if not already)
install.packages("devtools")

# Load the relevant library
library(devtools)

# Install the arsampler package from GitHub
devtools::install_github("enggelin/arsampler", INSTALL_opts = "--install-tests")

# Load the arsampler package
library(arsampler)
```


## Get started with arsampler

Read the short introductory vignette to get you started with `arsampler`. To look at examples, follow the section **Example** right after this section.

```{r getstarted}
# Read the short vignette
vignette("introduction-to-arsampler")
```

The vignette above basically explains that the package, in its current version, has 3 functions: `ar()`, which is the main function and outputs an object of class `ar`; `ar_summary()`, which receives input of an object of class `ar` and outputs the summary of the fitted densities; and `ar_plot()`, which receives input of an object of class `ar` and outputs the overlay plot of both the target and proposal densities (corrected with `c`, if defined) in a single plot. On this version, only two functions are ready to be used and tested, i.e. `ar()` and `ar_summary()`.


## Example

This is a basic example of the usage of `arsampler` package:

```{r example, warning=FALSE, message=FALSE}
library(arsampler)

# Target density: Normal(0,1); Proposal density: Normal(0,1)
f_norm <- "(1/sqrt(2*pi*1^2))*exp((-(x-0)^2)/(2*1^2))"
q_norm <- "rnorm(n, 0, 1)"

# Run the main function ar()
example_norm <- ar(f_norm, q_norm, c=1, n=10)
```

```{r, warning=FALSE, message=FALSE}
# Print the raw values of the fitted densities
example_norm
```

```{r, warning=FALSE, message=FALSE}
# Provide a summary of the fitted densities
ar_summary(example_norm)
```

The above output can also be achieved by using the commands `print()` and `summary()`, which is an S3 object-oriented method in R. Those commands are overridden if the user passes an object of class `ar`, which is the object class of the output from the function `ar()`.

```{r, warning=FALSE, message=FALSE}
# Print the raw values of the fitted densities using S3 method print()
print(example_norm)
```

```{r, warning=FALSE, message=FALSE}
# Provide a summary of the fitted densities using S3 method summary()
summary(example_norm)
```


## Test plan: unit testing

In this version, we will test several scenarios involving varieties of pairs of target densities and pre-defined proposal densities and how they will perform.

The unit tests can be run directly from the tests folder obtained when installing the package. There are one test file in the tests folder, which is used to test the main function of this package, `ar()`.

To run the unit tests from the tests folder, run the following command:

``` r
# Install the relevant library (if not already)
install.package("testthat")

# Load the relevant library
library(testthat)

# Set the working directory to the package root, and run this command:
testthat::test_dir(system.file("tests", package = "arsampler"))
```

The above command will test several scenarios defined by the author of this package. The unit testing comprises of several scenario which may arise in certain circumstances and the expected output from it. Inside the directory `tests/testthat/`, there is one test file names `test-ar.R`. The content of the file is as follows:

``` r
test_that("The test of the output for the exact same pair of target and proposal densities.", {

  ## initialise variables for testing
  # pair of standard normal distributions
  f_norm <- "(1/sqrt(2*pi*1^2))*exp((-(x-0)^2)/(2*1^2))"
  q_norm <- "rnorm(n, 0, 1)"

  # pair of uniform distribution
  f_unif <- "1"
  q_unif <- "runif(n, 0, 1)"

  # pair of gamma distribution, alpha=3, beta=2
  f_gamma <- "((2^3)/(gamma(3)))*x^(3-1)*exp(-2*x)"
  q_gamma <- "rgamma(n, 3, 2)"

  # pair of beta distribution, alpha=3, beta=2
  f_beta <- "((gamma(3+2))/gamma(3)*gamma(2))*x^(3-1)*(1-x)^(2-1)"
  q_beta <- "rbeta(n, 3, 2)"

  # pair of exp distribution, lambda=3
  f_exp <- "3*exp(-3*x)"
  q_exp <- "rexp(n, 3)"

  # expect silent (correct input)
  expect_silent(ar(f_norm, q_norm, c=1, n=1000))
  expect_silent(ar(f_unif, q_unif, c=1, n=1000))
  expect_silent(ar(f_gamma, q_gamma, c=1, n=1000))
  expect_silent(ar(f_beta, q_beta, c=1, n=1000))
  expect_silent(ar(f_exp, q_exp, c=1, n=1000))

  # expect equal: acceptance rate = 1
  expect_equal(ar(f_norm, q_norm, c=1, n=1000)$acceptance_rate, 1)
  expect_equal(ar(f_unif, q_unif, c=1, n=1000)$acceptance_rate, 1)
  expect_equal(ar(f_gamma, q_gamma, c=1, n=1000)$acceptance_rate, 1)
  expect_equal(ar(f_beta, q_beta, c=1, n=1000)$acceptance_rate, 1)
  expect_equal(ar(f_exp, q_exp, c=1, n=1000)$acceptance_rate, 1)

  # expect equal: rounded mean from standard distributions; longer run for convergence
  expect_equal(abs(round(ar(f_norm, q_norm, c=1, n=10000)$expected_values_f, 1)), 0) # need to put abs() on the standard normal distribution because the possible value may range from negative around zero to positive around zero.
  expect_equal(round(ar(f_unif, q_unif, c=1, n=10000)$expected_values_f, 1), 0.5)
  expect_equal(round(ar(f_gamma, q_gamma, c=1, n=10000)$expected_values_f, 1), 1.5)
  expect_equal(round(ar(f_beta, q_beta, c=1, n=10000)$expected_values_f, 1), 0.6)
  expect_equal(round(ar(f_exp, q_exp, c=1, n=10000)$expected_values_f, 1), 0.3)

  # expect match: suggested c = NA
  expect_match(ar(f_norm, q_norm, c=1, n=1000)$suggested_c, "NA")
  expect_match(ar(f_unif, q_unif, c=1, n=1000)$suggested_c, "NA")
  expect_match(ar(f_gamma, q_gamma, c=1, n=1000)$suggested_c, "NA")
  expect_match(ar(f_beta, q_beta, c=1, n=1000)$suggested_c, "NA")
  expect_match(ar(f_exp, q_exp, c=1, n=1000)$suggested_c, "NA")
})
```

If the test runs smoothly, the console with return `20` passed test item.

The content tested by the test file can be done by the user using the instruction below.


## Test plan: user input

Besides testing using the unit tests embedded in the installed package, user can also perform relevant testings by their own using their own input. The details are explain below.

You will notice that in the above example, the expected mean and variance are rather nonsense, as we only run the sampler for a few iterations. To be able to provide a rather accurate results, we need to run the sampler long enough. An iteration of minimum of `n=1000` should provide suffice (this is also the default value for `n`), although for some densities it needs to be run long enough (more about that later). In this test plan, we will try several scenarios of combination pair of target-proposal densities.


### Test 1: Proposal densities with Acceptance rate: 1, Suggested minimum c: NA, and Empirical expected value of target distribution $\approx$ its mean

The package has several pre-defined probability densities for user to use. They are: `normal`, `uniform`, `gamma`, `beta`, and `exponential`. To check whether this proposal densities give the correct output for its counterpart target densities written in its pdf form, you can run these tests below (notice that we will run it for longer iterations, i.e. `n=1000` and no parameters declaration are done as instructed in the vignette):

```{r test1}
# pair of standard normal distributions
f_norm <- "(1/sqrt(2*pi*1^2))*exp((-(x-0)^2)/(2*1^2))"
q_norm <- "rnorm(n, 0, 1)"

example_norm <- ar(f_norm, q_norm, c=1, n=1000)
summary(example_norm)
```

```{r test2}
# pair of uniform distributions
f_unif <- "1"
q_unif <- "runif(n, 0, 1)"

example_unif <- ar(f_unif, q_unif, c=1, n=1000)
summary(example_unif)
```

```{r test3}
# pair of gamma distributions, alpha=3, beta=2
f_gamma <- "((2^3)/(gamma(3)))*x^(3-1)*exp(-2*x)"
q_gamma <- "rgamma(n, 3, 2)"

example_gamma <- ar(f_gamma, q_gamma, c=1, n=1000)
summary(example_gamma)
```

```{r test4}
# pair of beta distributions, alpha=3, beta=2
f_beta <- "((gamma(3+2))/gamma(3)*gamma(2))*x^(3-1)*(1-x)^(2-1)"
q_beta <- "rbeta(n, 3, 2)"

example_beta <- ar(f_beta, q_beta, c=1, n=1000)
summary(example_beta)
```

```{r test5}
# pair of exp distributions, lambda=3
f_exp <- "3*exp(-3*x)"
q_exp <- "rexp(n, 3)"

example_exp <- ar(f_exp, q_exp, c=1, n=1000)
summary(example_exp)
```

Notice that for the above examples, we do not use the command `print()`, as we do not want to clutter this section. Another S3 method, `head()`, will be implemented in a later version to accommodate this shortcoming.

If the examples run correctly, you will notice that each printed summary gives `Acceptance rate: 1`, `Suggested minimum c: NA`, and `Empirical expected value of target distribution: (approximately its mean)`. This means that the proposal densities perfectly match their own target densities, as they are the exact same distributions with the exact same parameters. `Acceptance rate: 1` means all the proposed values are accepted as coming from the corresponding target distribution, `Suggested minimum c: NA` means that the proposal distribution envelopes the target distribution entirely with the given value of `c`, and `Empirical expected value of target distribution: (approximately its mean)` means that the accepted `x` values from the proposal distributions are able to approximate the mean of the target distribution. You can input your own parameters and see if the expected value match with the corresponding distributions. For reference:

Normal distribution with $\mu=0$ and $\sigma^2=1$: $X \sim N(0,1)$; $E(X)=\mu$, $Var(X)=\sigma^2$;

Uniform distribution with $a=0$ and $b=1$: $X \sim U(0,1)$; $E(X)=\frac{a+b}{2}$, $Var(X)=\frac{(b-a)^2}{12}$;

Gamma distribution with $\alpha=3$ and $\beta=2$: $X \sim Gamma(3,2)$; $E(X)=\frac{\alpha}{\beta}$, $Var(X)=\frac{\alpha}{\beta^2}$;

Beta distribution with $\alpha=3$ and $\beta=2$: $X \sim Beta(3,2)$; $E(X)=\frac{\alpha}{\alpha+\beta}$, $Var(X)=\frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$;

Exponential distribution with $\lambda=3$: $X \sim Exp(3)$; $E(X)=\frac{1}{\lambda}$, $Var(X)=\frac{1}{\lambda^2}$;

You can also try different parameters to see how it works.


### Test 2: Proposal densities with different form than the target densities

In the real world usage, the case is not that simple; most of the time we need to estimate the target distribution from a different, usually easier, proposal distribution (which also coincides with the purpose of this package). We can test several simple pair of target-proposal densities, as instructed below:

```{r test6}
# pair of standard normal density (f) and uniform density (q)
f_norm <- "(1/sqrt(2*pi*1^2))*exp((-(x-0)^2)/(2*1^2))"
q_unif <- "runif(n, -4, 4)" # to accommodate the lower and upper limit of N(0,1) distribution

example_norm_unif <- ar(f_norm, q_unif, c=1, n=10000) # run longer for convergence
summary(example_norm_unif)
```

From the result above, we can see that the proposal density does not fully envelope the target density, determined by the value of `Suggested minimum c`. Also, the expected value and variance are nonsense. We will follow the suggested value of `c` and re-run the example code:

```{r test7}
# pair of standard normal density (f) and uniform density (q)
f_norm <- "(1/sqrt(2*pi*1^2))*exp((-(x-0)^2)/(2*1^2))"
q_unif <- "runif(n, -4, 4)" # to accommodate the lower and upper limit of N(0,1) distribution

example_norm_unif <- ar(f_norm, q_unif, c=3.2, n=10000) # change the value of c to the suggested minimum.
summary(example_norm_unif)
```

The suggested minimum `c` is now `NA`, which means the proposal density now envelopes the target density fully, and the expected value and variance are now valid. ($\hat{E(X)} \approx \mu$ and $\hat{Var(X)} \approx \sigma^2$).

You can try using different combinations of target-density pair for this part.
